---
title: "Complete Tutorial"
subtitle: "End-to-End Workflow for Scientific Content Analysis"
---

## Introduction

This tutorial provides a complete workflow for analyzing scientific papers using the contentanalysis package. We'll work through a real example, from PDF import to final visualizations and reporting.

## Setup

### Install Required Packages

```{r setup, eval=FALSE}
# Install contentanalysis
devtools::install_github("massimoaria/contentanalysis")

# Install supporting packages
install.packages(c("dplyr", "ggplot2", "tidyr", "knitr"))

# Load libraries
library(contentanalysis)
library(dplyr)
library(ggplot2)
library(tidyr)
```

::: {.callout-tip}
## Optional: Setup AI-Enhanced Features 🆕

For improved PDF extraction with complex layouts:

```{r setup_ai, eval=FALSE}
# Get API key from https://aistudio.google.com/apikey
Sys.setenv(GEMINI_API_KEY = "your-api-key-here")

# Or add to .Renviron file:
# GEMINI_API_KEY=your-api-key-here
```
:::

## Step 1: Obtain Sample Paper

We'll use an open-access paper on Machine Learning:

```{r download, eval=FALSE}
# Download example paper
paper_url <- "https://raw.githubusercontent.com/massimoaria/contentanalysis/master/inst/examples/example_paper.pdf"
download.file(paper_url, destfile = "example_paper.pdf", mode = "wb")

# Verify download
file.exists("example_paper.pdf")
```

::: {.callout-note}
## Using Your Own Papers

Replace the URL with your own PDF file path. Ensure the PDF is text-based (not a scanned image).
:::

## Step 2: Import and Inspect PDF

### Import with Section Detection

```{r import, eval=FALSE}
# Import PDF with automatic section detection
doc <- pdf2txt_auto(
  "example_paper.pdf",
  n_columns = 2,          # Two-column layout
  sections = TRUE         # Detect sections
)

# Check detected sections
cat("Detected sections:\n")
print(names(doc))

# Preview Abstract
cat("\n=== Abstract Preview ===\n")
cat(substr(doc$Abstract, 1, 500), "...\n")
```

### Verify Section Quality

```{r verify, eval=FALSE}
# Check section word counts
section_lengths <- sapply(doc[names(doc) != "Full_text"], function(x) {
  length(strsplit(x, "\\s+")[[1]])
})

section_df <- data.frame(
  section = names(section_lengths),
  words = section_lengths
) %>%
  arrange(desc(words))

print(section_df)

# Visualize section lengths
ggplot(section_df, aes(x = reorder(section, words), y = words, fill = section)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Word Count by Section",
       x = "Section", y = "Number of Words") +
  theme_minimal()
```

## Step 3: Comprehensive Content Analysis

### Main Analysis

```{r analysis, eval=FALSE}
# Perform comprehensive analysis with enhanced metadata integration
analysis <- analyze_scientific_content(
  text = doc,
  doi = "10.1016/j.mlwa.2021.100094",  # Paper's DOI
  mailto = "your@email.com",            # Your email for CrossRef
  window_size = 10,                     # Context window
  remove_stopwords = TRUE,              # Remove common words
  ngram_range = c(1, 3),               # Unigrams to trigrams
  use_sections_for_citations = TRUE
)

# View summary
print(analysis$summary)
```

::: {.callout-note}
## 🆕 Enhanced Features

The analysis now includes:

- **Dual metadata integration**: Automatically retrieves references from both CrossRef and OpenAlex
- **Improved citation matching**: Better handling of numeric citations (`[1]`, `[1-3]`) and author-year formats
- **Enhanced confidence scoring**: More granular assessment of match quality
- **Better author name handling**: Resolves variants like "Smith, J." vs "Smith, John"
:::

### Interpret Summary Statistics

```{r interpret, eval=FALSE}
# Extract key metrics
total_words <- analysis$summary$total_words
citations <- analysis$summary$citations_extracted
density <- analysis$summary$citation_density
diversity <- analysis$summary$lexical_diversity

cat("Document Statistics:\n")
cat("===================\n")
cat(sprintf("Total words: %d\n", total_words))
cat(sprintf("Citations: %d\n", citations))
cat(sprintf("Citation density: %.2f per 1000 words\n", density))
cat(sprintf("Lexical diversity: %.3f\n", diversity))

# Assess citation intensity
if (density < 5) {
  cat("\n→ Low citation density (typical for theoretical papers)\n")
} else if (density < 15) {
  cat("\n→ Moderate citation density (standard empirical paper)\n")
} else {
  cat("\n→ High citation density (review paper or methods paper)\n")
}
```

## Step 4: Citation Analysis

### Extract and Explore Citations

```{r citations, eval=FALSE}
# View first citations
head(analysis$citations, 10)

# Citation types
citation_summary <- analysis$citations %>%
  group_by(citation_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(percentage = round(count / sum(count) * 100, 1))

print(citation_summary)

# Visualize
ggplot(citation_summary, aes(x = citation_type, y = count, fill = citation_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(count, " (", percentage, "%)")), 
            vjust = -0.5) +
  labs(title = "Citation Types",
       x = "Type", y = "Count") +
  theme_minimal()
```

### Citations by Section

```{r sections_cit, eval=FALSE}
# Citation distribution across sections
section_citations <- analysis$citations %>%
  count(section, sort = TRUE)

print(section_citations)

# Visualize
ggplot(section_citations, aes(x = reorder(section, n), y = n, fill = section)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Citations by Section",
       x = "Section", y = "Number of Citations") +
  theme_minimal()
```

### Most Cited References

```{r most_cited, eval=FALSE}
# Top 10 most cited references
top_cited <- analysis$citation_references_mapping %>%
  count(ref_full_text, sort = TRUE) %>%
  head(10) %>%
  mutate(ref_short = substr(ref_full_text, 1, 60))

print(top_cited)

# Visualize
ggplot(top_cited, aes(x = reorder(ref_short, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Most Cited References",
       x = NULL, y = "Citation Count") +
  theme_minimal()
```

### Citation Contexts

```{r contexts, eval=FALSE}
# Examine citation contexts
contexts_sample <- analysis$citation_contexts %>%
  select(citation_text_clean, section, words_before, words_after) %>%
  head(5)

print(contexts_sample)

# Find method citations
method_citations <- analysis$citation_contexts %>%
  filter(grepl("method|approach|algorithm|technique", 
               paste(words_before, words_after), 
               ignore.case = TRUE)) %>%
  select(citation_text_clean, section, words_before, words_after)

cat("\nMethod-related citations found:", nrow(method_citations), "\n")
head(method_citations)
```

## Step 5: Network Visualization

### Create Citation Network

```{r network, eval=FALSE}
# Create interactive network
network <- create_citation_network(
  citation_analysis_results = analysis,
  max_distance = 800,
  min_connections = 2,
  show_labels = TRUE
)

# Display network
network
```

### Analyze Network Statistics

```{r network_stats, eval=FALSE}
# Get network statistics
stats <- attr(network, "stats")

cat("Network Statistics:\n")
cat("===================\n")
cat("Nodes:", stats$n_nodes, "\n")
cat("Edges:", stats$n_edges, "\n")
cat("Avg distance:", round(stats$avg_distance), "characters\n")
cat("Max distance:", stats$max_distance, "characters\n")

# Network density
density <- stats$n_edges / (stats$n_nodes * (stats$n_nodes - 1) / 2)
cat("Network density:", round(density, 3), "\n")

# Section distribution
print(stats$section_distribution)

# Hub citations
hub_threshold <- quantile(stats$section_distribution$n, 0.75)
hubs <- stats$section_distribution %>%
  filter(n >= hub_threshold) %>%
  arrange(desc(n))

cat("\nHub citations (top 25%):\n")
print(hubs)
```

### Export Network

```{r export_network, eval=FALSE}
library(htmlwidgets)

# Save as standalone HTML
saveWidget(network, 
           "citation_network.html",
           selfcontained = TRUE,
           title = "Citation Network")

cat("Network saved to: citation_network.html\n")
```

## Step 6: Text Analysis

### Word Frequency Analysis

```{r word_freq, eval=FALSE}
# Top 30 words
top_words <- head(analysis$word_frequencies, 30)
print(top_words)

# Visualize top 20
top_20 <- head(analysis$word_frequencies, 20)

ggplot(top_20, aes(x = reorder(word, frequency), y = frequency)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(title = "Top 20 Most Frequent Words",
       x = "Word", y = "Frequency") +
  theme_minimal()
```

### N-gram Analysis

```{r ngrams, eval=FALSE}
# Most common bigrams
top_bigrams <- head(analysis$ngrams$`2gram`, 15)
print(top_bigrams)

# Most common trigrams
top_trigrams <- head(analysis$ngrams$`3gram`, 10)
print(top_trigrams)

# Visualize bigrams
ggplot(top_bigrams, aes(x = reorder(ngram, frequency), y = frequency)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(title = "Top 15 Bigrams",
       x = "Bigram", y = "Frequency") +
  theme_minimal()
```

### Word Distribution Tracking

```{r distribution, eval=FALSE}
# Define key terms to track
key_terms <- c("machine learning", "random forest", "accuracy", 
               "classification", "model")

# Calculate distribution
dist <- calculate_word_distribution(
  text = doc,
  selected_words = key_terms,
  use_sections = TRUE,
  normalize = TRUE
)

# View results
print(dist)

# Interactive visualization
plot_word_distribution(
  dist,
  plot_type = "line",
  show_points = TRUE,
  smooth = TRUE
)
```

## Step 7: Readability Assessment

### Overall Readability

```{r readability, eval=FALSE}
# Calculate readability for full text
readability <- calculate_readability_indices(
  doc$Full_text,
  detailed = TRUE
)

print(readability)

# Interpret
cat("\nInterpretation:\n")
cat("Flesch Reading Ease:", readability$flesch_reading_ease, "\n")
if (readability$flesch_reading_ease < 30) {
  cat("→ Very difficult (graduate level)\n")
} else if (readability$flesch_reading_ease < 50) {
  cat("→ Difficult (college level)\n")
} else {
  cat("→ Fairly difficult (high school to college)\n")
}

cat("\nGrade Level:", round(readability$flesch_kincaid_grade, 1), "\n")
```

### Compare Sections

```{r readability_sections, eval=FALSE}
# Calculate for each section
sections <- c("Abstract", "Introduction", "Methods", "Results", "Discussion")
section_readability <- data.frame()

for (section in sections) {
  if (section %in% names(doc)) {
    metrics <- calculate_readability_indices(doc[[section]], detailed = FALSE)
    metrics$section <- section
    section_readability <- rbind(section_readability, metrics)
  }
}

print(section_readability)

# Visualize
ggplot(section_readability, 
       aes(x = reorder(section, flesch_reading_ease), 
           y = flesch_reading_ease, fill = section)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Readability by Section",
       subtitle = "Higher scores = easier to read",
       x = "Section", y = "Flesch Reading Ease") +
  theme_minimal()
```

## Step 8: Comprehensive Reporting

### Create Summary Report

```{r report, eval=FALSE}
# Compile comprehensive report
report <- list(
  document_info = list(
    doi = "10.1016/j.mlwa.2021.100094",
    total_words = analysis$summary$total_words,
    sections = names(doc)[names(doc) != "Full_text"]
  ),
  
  citation_metrics = list(
    total_citations = analysis$summary$citations_extracted,
    narrative = analysis$summary$narrative_citations,
    parenthetical = analysis$summary$parenthetical_citations,
    matched = analysis$summary$references_matched,
    density = analysis$summary$citation_density
  ),
  
  text_metrics = list(
    lexical_diversity = analysis$summary$lexical_diversity,
    top_10_words = head(analysis$word_frequencies$word, 10),
    top_10_bigrams = head(analysis$ngrams$`2gram`$ngram, 10)
  ),
  
  readability = list(
    flesch_reading_ease = readability$flesch_reading_ease,
    grade_level = readability$flesch_kincaid_grade,
    gunning_fog = readability$gunning_fog
  ),
  
  network_stats = list(
    nodes = stats$n_nodes,
    edges = stats$n_edges,
    density = stats$n_edges / (stats$n_nodes * (stats$n_nodes - 1) / 2)
  )
)

# Print report
cat("COMPREHENSIVE ANALYSIS REPORT\n")
cat("=============================\n\n")

cat("DOCUMENT INFORMATION\n")
cat("DOI:", report$document_info$doi, "\n")
cat("Total words:", report$document_info$total_words, "\n")
cat("Sections:", paste(report$document_info$sections, collapse = ", "), "\n\n")

cat("CITATION METRICS\n")
cat("Total citations:", report$citation_metrics$total_citations, "\n")
cat("Citation density:", round(report$citation_metrics$density, 2), "per 1000 words\n")
cat("Match rate:", round(report$citation_metrics$matched / report$citation_metrics$total_citations * 100, 1), "%\n\n")

cat("TEXT METRICS\n")
cat("Lexical diversity:", round(report$text_metrics$lexical_diversity, 3), "\n")
cat("Top words:", paste(head(report$text_metrics$top_10_words, 5), collapse = ", "), "\n\n")

cat("READABILITY\n")
cat("Reading ease:", round(report$readability$flesch_reading_ease, 1), "\n")
cat("Grade level:", round(report$readability$grade_level, 1), "\n\n")

cat("NETWORK STATISTICS\n")
cat("Citation nodes:", report$network_stats$nodes, "\n")
cat("Connections:", report$network_stats$edges, "\n")
cat("Density:", round(report$network_stats$density, 3), "\n")
```

### Export All Results

```{r export_all, eval=FALSE}
# Create output directory
dir.create("analysis_output", showWarnings = FALSE)

# 1. Citations
write.csv(analysis$citations, 
          "analysis_output/citations.csv", 
          row.names = FALSE)

# 2. Matched references
write.csv(analysis$citation_references_mapping,
          "analysis_output/matched_references.csv",
          row.names = FALSE)

# 3. Word frequencies
write.csv(analysis$word_frequencies,
          "analysis_output/word_frequencies.csv",
          row.names = FALSE)

# 4. Bigrams
write.csv(analysis$ngrams$`2gram`,
          "analysis_output/bigrams.csv",
          row.names = FALSE)

# 5. Trigrams
write.csv(analysis$ngrams$`3gram`,
          "analysis_output/trigrams.csv",
          row.names = FALSE)

# 6. Network statistics
write.csv(stats$section_distribution,
          "analysis_output/network_stats.csv",
          row.names = FALSE)

# 7. Readability by section
write.csv(section_readability,
          "analysis_output/readability.csv",
          row.names = FALSE)

# 8. Summary report as JSON
library(jsonlite)
write_json(report, 
           "analysis_output/summary_report.json",
           pretty = TRUE, 
           auto_unbox = TRUE)

cat("All results exported to: analysis_output/\n")
```

## Step 9: Advanced Visualizations

### Create Publication-Ready Figures

```{r pub_figs, eval=FALSE}
library(patchwork)

# Figure 1: Overview
p1 <- ggplot(section_df, aes(x = reorder(section, words), y = words)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "A) Document Structure", x = NULL, y = "Words") +
  theme_minimal()

p2 <- ggplot(citation_summary, aes(x = citation_type, y = count, fill = citation_type)) +
  geom_col(show.legend = FALSE) +
  labs(title = "B) Citation Types", x = NULL, y = "Count") +
  theme_minimal()

# Combine
combined <- p1 + p2
print(combined)

ggsave("analysis_output/figure1_overview.png", 
       combined, width = 10, height = 5, dpi = 300)

# Figure 2: Text analysis
p3 <- ggplot(head(top_words, 15), 
             aes(x = reorder(word, frequency), y = frequency)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(title = "A) Top Words", x = NULL, y = "Frequency") +
  theme_minimal()

p4 <- ggplot(section_readability, 
             aes(x = reorder(section, flesch_reading_ease), 
                 y = flesch_reading_ease)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(title = "B) Readability", x = NULL, y = "FRE Score") +
  theme_minimal()

combined2 <- p3 + p4
print(combined2)

ggsave("analysis_output/figure2_text_analysis.png",
       combined2, width = 10, height = 5, dpi = 300)
```

## Step 10: Batch Processing

### Analyze Multiple Papers

```{r batch, eval=FALSE}
# Define papers to analyze
papers_df <- data.frame(
  file = c("paper1.pdf", "paper2.pdf", "paper3.pdf"),
  doi = c("10.xxxx/1", "10.xxxx/2", "10.xxxx/3"),
  name = c("Paper A", "Paper B", "Paper C"),
  stringsAsFactors = FALSE
)

# Process all papers
all_results <- list()
all_networks <- list()

for (i in 1:nrow(papers_df)) {
  cat("\nProcessing:", papers_df$name[i], "\n")
  
  # Import
  doc <- pdf2txt_auto(papers_df$file[i], n_columns = 2)
  
  # Analyze
  all_results[[i]] <- analyze_scientific_content(
    text = doc,
    doi = papers_df$doi[i],
    mailto = "your@email.com"
  )
  
  # Network
  all_networks[[i]] <- create_citation_network(
    all_results[[i]],
    max_distance = 800,
    min_connections = 2
  )
  
  Sys.sleep(1)  # Be polite to CrossRef API
}

names(all_results) <- papers_df$name
names(all_networks) <- papers_df$name

# Compare papers
comparison <- data.frame(
  paper = papers_df$name,
  words = sapply(all_results, function(r) r$summary$total_words),
  citations = sapply(all_results, function(r) r$summary$citations_extracted),
  density = sapply(all_results, function(r) r$summary$citation_density),
  diversity = sapply(all_results, function(r) r$summary$lexical_diversity),
  network_nodes = sapply(all_networks, function(n) attr(n, "stats")$n_nodes),
  network_edges = sapply(all_networks, function(n) attr(n, "stats")$n_edges)
)

print(comparison)

# Visualize comparison
comparison_long <- comparison %>%
  select(paper, citations, density, diversity) %>%
  pivot_longer(cols = -paper, names_to = "metric", values_to = "value")

ggplot(comparison_long, aes(x = paper, y = value, fill = paper)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "Comparison Across Papers",
       x = NULL, y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Conclusion

This tutorial covered the complete workflow:

1. ✓ PDF import with section detection
2. ✓ Comprehensive content analysis
3. ✓ Citation extraction and matching
4. ✓ Interactive network visualization
5. ✓ Text analysis and n-grams
6. ✓ Readability assessment
7. ✓ Comprehensive reporting
8. ✓ Data export
9. ✓ Publication-ready figures
10. ✓ Batch processing

## Next Steps

- Explore [Reference Documentation](reference/) for detailed function information
- Try the analysis on your own papers
- Customize visualizations for your needs
- Integrate into your research workflow

## Resources

- [GitHub Repository](https://github.com/massimoaria/contentanalysis)
- [Issue Tracker](https://github.com/massimoaria/contentanalysis/issues)
- [Get Started Guide](get-started.qmd)

## Troubleshooting

### Common Issues

**PDF Import Problems**
```r
# Try different column settings
doc1 <- pdf2txt_auto("paper.pdf", n_columns = 1)
doc2 <- pdf2txt_auto("paper.pdf", n_columns = 2)
# Compare which works better
```

**Low Citation Matching**
```r
# Ensure DOI and email are provided
# Check References section was extracted
names(doc)  # Should include "References"
```

**Network Not Displaying**
```r
# Adjust parameters
network <- create_citation_network(
  analysis,
  max_distance = 1000,  # Increase
  min_connections = 1    # Decrease
)
```

For more help, see the [Get Started](get-started.qmd) troubleshooting section.