---
title: "PDF Import Functions"
---

## Overview

The PDF import functions handle the extraction of text from PDF documents while preserving document structure. The package provides automatic section detection, supports various column layouts, and now includes **AI-enhanced extraction** through Google's Gemini API for handling complex document layouts.

::: {.callout-note}
## 🆕 AI-Enhanced Extraction

The package now supports AI-assisted PDF text extraction for improved accuracy with complex layouts. This feature uses Google's Gemini API to better parse:

- Multi-column layouts with complex flow
- Documents with tables and figures
- Papers with unusual formatting
- Scanned or low-quality PDFs

To use AI features, you need a Gemini API key from [Google AI Studio](https://aistudio.google.com/apikey).
:::

## Main Function

### pdf2txt_auto()

Import PDF with automatic section detection and multi-column support. Now includes optional AI enhancement.

**Usage**

```r
pdf2txt_auto(
  file_path,
  n_columns = 1,
  sections = TRUE,
  section_keywords = NULL,
  use_ai = FALSE,        # New: Enable AI processing
  ai_model = "2.0-flash" # New: Gemini model version
)
```

**Arguments**

- `file_path`: Character string. Path to the PDF file
- `n_columns`: Integer. Number of columns in the PDF layout (1, 2, or 3)
- `sections`: Logical. Whether to split text into sections
- `section_keywords`: Character vector. Custom keywords for section detection
- `use_ai`: Logical. Whether to use AI-enhanced extraction (default: FALSE) 🆕
- `ai_model`: Character. Gemini model version to use (default: "2.0-flash") 🆕

**Value**

A named list containing:

- `Full_text`: The complete document text
- Section-specific text (e.g., `Abstract`, `Introduction`, `Methods`, etc.)

## New AI Functions 🆕

### gemini_content_ai()

Process PDFs and images with Google Gemini AI for enhanced content extraction.

**Usage**

```r
gemini_content_ai(
  image = NULL,
  docs = NULL,
  prompt = "Explain these images",
  model = "2.0-flash",
  image_type = "png",
  retry_503 = 5,
  api_key = NULL,
  outputSize = "medium"
)
```

**Arguments**

- `image`: Character vector. Path(s) to image file(s)
- `docs`: Character vector. Path(s) to document file(s) (PDF, TXT, HTML, CSV, RTF)
- `prompt`: Character. The prompt/instruction for the AI model
- `model`: Character. Gemini model version ("1.5-flash", "2.0-flash", "2.5-flash")
- `image_type`: Character. Image MIME type (default: "png")
- `retry_503`: Integer. Retry attempts for HTTP 503 errors (default: 5)
- `api_key`: Character. Gemini API key (uses GEMINI_API_KEY env var if NULL)
- `outputSize`: Character. Output token limit: "small" (8K), "medium" (16K), "large" (32K), "huge" (131K)

**Value**

Character vector containing the AI-generated response(s).

### process_large_pdf() 🆕

Process large PDFs in chunks using AI for better handling of lengthy documents.

**Usage**

```r
process_large_pdf(
  file_path,
  chunk_pages = 10,
  ai_model = "2.0-flash",
  prompt = "Extract and structure the text from this document",
  api_key = NULL
)
```

**Arguments**

- `file_path`: Character string. Path to the PDF file
- `chunk_pages`: Integer. Number of pages to process per chunk (default: 10)
- `ai_model`: Character. Gemini model version to use
- `prompt`: Character. Instruction for the AI model
- `api_key`: Character. Gemini API key

**Value**

A named list with text extracted from all chunks, merged together.

## Examples

### Basic Import

Import a simple single-column PDF:

```{r basic, eval=FALSE}
library(contentanalysis)

# Single column PDF
doc <- pdf2txt_auto("paper.pdf", n_columns = 1)

# Check structure
names(doc)
str(doc, max.level = 1)
```

### Multi-Column PDFs

Most academic papers use two-column layouts:

```{r multicolumn, eval=FALSE}
# Two-column layout (most common)
doc <- pdf2txt_auto("paper.pdf", n_columns = 2)

# Three-column layout (less common)
doc_three <- pdf2txt_auto("paper.pdf", n_columns = 3)
```

### Section Detection

The function automatically detects common academic sections:

```{r sections, eval=FALSE}
# With automatic section detection (default)
doc <- pdf2txt_auto("paper.pdf", n_columns = 2, sections = TRUE)

# View detected sections
names(doc)
# [1] "Full_text"    "Abstract"     "Introduction" "Methods"     
# [5] "Results"      "Discussion"   "References"

# Access specific sections
cat(doc$Abstract)
cat(doc$Introduction)
```

### Without Section Splitting

Extract text without section detection:

```{r no_sections, eval=FALSE}
# Get only full text
text_only <- pdf2txt_auto("paper.pdf", sections = FALSE)

# Result is a list with only Full_text
names(text_only)
# [1] "Full_text"
```

### AI-Enhanced Import 🆕

Use AI for complex or difficult PDFs:

```{r ai_enhanced, eval=FALSE}
# Set API key (if not in .Renviron)
Sys.setenv(GEMINI_API_KEY = "your-api-key-here")

# Basic AI-enhanced extraction
doc_ai <- pdf2txt_auto(
  "complex_paper.pdf",
  n_columns = 2,
  use_ai = TRUE,
  ai_model = "2.0-flash"
)

# Process a large PDF in chunks
large_doc <- process_large_pdf(
  "long_paper.pdf",
  chunk_pages = 10,
  ai_model = "2.0-flash"
)

# Direct AI extraction with custom prompt
result <- gemini_content_ai(
  docs = "paper.pdf",
  prompt = "Extract all section headings and their content from this paper",
  outputSize = "large"
)

# Process multiple documents
results <- gemini_content_ai(
  docs = c("paper1.pdf", "paper2.pdf"),
  prompt = "Summarize the main findings",
  model = "2.0-flash"
)
```

### Custom Section Keywords

Define custom keywords for section detection:

```{r custom, eval=FALSE}
# Custom section keywords
my_keywords <- list(
  Background = c("background", "literature review"),
  Methodology = c("methodology", "experimental design"),
  Findings = c("findings", "observations"),
  Conclusions = c("conclusions", "summary")
)

doc <- pdf2txt_auto(
  "paper.pdf",
  n_columns = 2,
  sections = TRUE,
  section_keywords = my_keywords
)

names(doc)
```

## Working with Results

### Accessing Section Content

```{r access, eval=FALSE}
# Get word count per section
section_lengths <- sapply(doc[names(doc) != "Full_text"], function(x) {
  length(strsplit(x, "\\s+")[[1]])
})

print(section_lengths)

# Example output:
#    Abstract Introduction      Methods      Results   Discussion   References 
#         234         1052          876         1342          945          523
```

### Checking Section Quality

```{r quality, eval=FALSE}
# Check if important sections were detected
required_sections <- c("Abstract", "Introduction", "Methods", "Results")
detected <- required_sections %in% names(doc)
names(detected) <- required_sections

print(detected)

# Identify missing sections
missing <- required_sections[!detected]
if (length(missing) > 0) {
  cat("Warning: Missing sections:", paste(missing, collapse = ", "), "\n")
}
```

### Preview Section Content

```{r preview, eval=FALSE}
# Preview first 500 characters of each section
preview_sections <- function(doc, n_chars = 500) {
  sections <- names(doc)[names(doc) != "Full_text"]
  
  for (section in sections) {
    cat("\n===", section, "===\n")
    cat(substr(doc[[section]], 1, n_chars), "...\n")
  }
}

preview_sections(doc)
```

## Common Use Cases

### Use Case 1: Prepare for Analysis

```{r usecase1, eval=FALSE}
# Import PDF
doc <- pdf2txt_auto("example_paper.pdf", n_columns = 2)

# Verify sections were detected
if (!"Introduction" %in% names(doc)) {
  warning("Introduction not detected - check section keywords")
}

# Pass to analysis
analysis <- analyze_scientific_content(
  text = doc,
  doi = "10.xxxx/xxxxx",
  mailto = "your@email.com"
)
```

### Use Case 2: Extract Specific Sections

```{r usecase2, eval=FALSE}
# Import multiple papers and extract only Methods
papers <- c("paper1.pdf", "paper2.pdf", "paper3.pdf")

methods_sections <- lapply(papers, function(p) {
  doc <- pdf2txt_auto(p, n_columns = 2)
  doc$Methods
})

names(methods_sections) <- papers

# Analyze methods section lengths
methods_lengths <- sapply(methods_sections, function(m) {
  if (!is.null(m)) {
    length(strsplit(m, "\\s+")[[1]])
  } else {
    NA
  }
})

print(methods_lengths)
```

### Use Case 3: Batch Processing

```{r usecase3, eval=FALSE}
# Process multiple papers with different layouts
process_paper <- function(file, columns = 2) {
  tryCatch({
    doc <- pdf2txt_auto(file, n_columns = columns)
    list(
      success = TRUE,
      sections = names(doc),
      word_count = length(strsplit(doc$Full_text, "\\s+")[[1]]),
      doc = doc
    )
  }, error = function(e) {
    list(
      success = FALSE,
      error = e$message,
      doc = NULL
    )
  })
}

# Process all PDFs in directory
pdf_files <- list.files(pattern = "\\.pdf$")
results <- lapply(pdf_files, process_paper)
names(results) <- pdf_files

# Check success rate
success_rate <- mean(sapply(results, function(r) r$success))
cat("Successfully processed:", success_rate * 100, "%\n")
```

## Tips and Best Practices

::: {.callout-tip}
## Column Detection

- **Start with n_columns = 2**: Most academic papers use two columns
- If text appears jumbled, try different column values
- Single-column: theses, reports, preprints
- Two-column: journal articles, conference papers
- Three-column: some conference proceedings
:::

::: {.callout-tip}
## Section Detection

- Works best with clearly formatted section headers
- Case-insensitive matching
- Handles common variations (e.g., "METHODS", "Methods", "Methodology")
- Custom keywords useful for non-standard formats
:::

::: {.callout-warning}
## Limitations

- Requires text-based PDFs (not scanned images)
- Complex layouts may affect text extraction
- Tables and figures may disrupt text flow
- Some formatting is lost in plain text conversion
:::

## Troubleshooting

### Problem: Garbled or Jumbled Text

**Solution**: Adjust the `n_columns` parameter

```{r troubleshoot1, eval=FALSE}
# Try different column settings
doc_1col <- pdf2txt_auto("paper.pdf", n_columns = 1)
doc_2col <- pdf2txt_auto("paper.pdf", n_columns = 2)
doc_3col <- pdf2txt_auto("paper.pdf", n_columns = 3)

# Compare results
sapply(list(doc_1col, doc_2col, doc_3col), function(d) {
  substr(d$Full_text, 1, 200)
})
```

### Problem: Sections Not Detected

**Solution**: Check section headers and use custom keywords

```{r troubleshoot2, eval=FALSE}
# Import without sections to see raw text
doc_raw <- pdf2txt_auto("paper.pdf", sections = FALSE)

# Search for section markers
section_markers <- c("abstract", "introduction", "methods", 
                     "results", "discussion", "conclusion")

positions <- sapply(section_markers, function(marker) {
  gregexpr(marker, doc_raw$Full_text, ignore.case = TRUE)[[1]][1]
})

print(positions[positions > 0])
```

### Problem: Missing References Section

**Solution**: References may be in a separate section or differently named

```{r troubleshoot3, eval=FALSE}
# Check all detected sections
names(doc)

# References might be under different names
possible_refs <- c("References", "Bibliography", "Works Cited", "Literature")
ref_section <- intersect(names(doc), possible_refs)

if (length(ref_section) > 0) {
  references <- doc[[ref_section]]
} else {
  cat("References section not found automatically\n")
}
```

## See Also

- [Content Analysis](content-analysis.qmd): Analyze imported documents
- [Citation Analysis](citation-analysis.qmd): Extract citations from text
- [Tutorial](../tutorial.qmd): Complete workflow examples